---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# peas

The purpose of this R package is to predict offspring distributions in genetic crosses. The package contains functions for defining systems of Mendelian inheritance (defined broadly),
and making predictions about the distribution of offspring genotypes and phenotypes
in genetic crosses. The package can handle arbitrary numbers of linkage groups, loci and alleles per locus. Several traits can be defined, potentially involving pleiotropy, polygeny and epistasis.

## Example

Here is a simple example for how one can define a genetic system along with a trait:
```{r}
library(peas)
MendelsPeas <- newGenopheno(alleleNames = list(c('Y', 'y')))
MendelsPeas <- setPhenotypes(MendelsPeas, 'colour', 'Y_', 'yellow')
MendelsPeas <- setPhenotypes(MendelsPeas, 'colour', 'yy', 'green')
MendelsPeas
```

Once defined, it is easy to predict offspring genotype and phenotype distributions in crosses:
```{r}
predictCross(MendelsPeas, 'Yy', 'Yy')
```

A detailed description of the package with more examples can be found in the package vignette (after installing the package, call `browseVignettes("peas")`).

## Installation

To install the peas package, you can run the following two lines of code:

```{r installation, eval = FALSE}
install.packages("devtools")
devtools::install_github("JanEngelstaedter/peas")
```
